---
title: "Chase's wheat project"
author: "Riley M. Anderson & Chase Baerlocher"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
graphics: yes
output:
  github_document:
    toc: yes
    toc_depth: 5
    pandoc_args: --webtex
  html_document:
    keep_md: yes
    theme: readable
    mathjax: default
  html_notebook:
    code_folding: hide
    theme: readable
    mathjax: default
  pdf_document:
    toc: yes
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
editor_options:
  chunk_output_type: console
---

```{r setup, include = F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set root directory to the project directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
#  set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))

```



## Overview

What is this analysis about?



```{r Main_Code, include = F, cache = F}

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Setup - This code is run, but output is hidden
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Load Packages
library(tidyverse) # Needed for data wrangling: dplyr, tidyr, ggplot2
library(car)
library(emmeans)

# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Data Preparation
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Import datasets
#data <- read.csv("data/file.csv")

pb <- read.csv("phys_bin.csv")
nb <- read.csv("nymph_bin.csv")
nodb <-read.csv("nod_bin.csv")

```

## Shoot weight

```{r Shoot_weight, echo = F, comment = ""}

pb$var<-factor(pb$var,levels = c("me","w","l","mi","d","k"),
               labels = c("Melrose","Windham","Lynx",
                          "MiCa","Dint","Klondike"))

pb$aph<-factor(pb$aph,levels = c(0,1,2),
               labels = c("None","PEMV-","PEMV+"))

pb$rhiz<-factor(pb$rhiz,levels = c(1,0),
                labels = c("R","NR"))

# ggplot(pb,aes(x=sw))+
#   geom_histogram(binwidth = 0.1, fill = "cyan", color = "black")+
#   labs(title = "SW", x = "Shoot Weight", y = "Count")+
#   theme_minimal()

# qqnorm(pb$sw)
# qqline(pb$sw,col="red")

# shapiro.test((pb$sw))
# we are interested in the normality of model residuals, not the raw data.

leveneTest(sw~var*rhiz*aph, data=pb)

sw.lm<-lm(sw~var*aph*rhiz,data = pb)

anova(sw.lm)
# The 3-way interaction is not significant, 
#      can we drop it and improve model fit?





sw.lm2way <- update(sw.lm, ~ . -(var:aph:rhiz))
# update the model to remove 3way ^

# likelihood ratio test (always compared to the full model, so type II SS)
anova(sw.lm, sw.lm2way, test = "Chisq")
# okay so removing the 3way doesn't make the model much better.

# v this is the same as sw.lm, they are identical models
sw.anova<-aov(sw~var*aph*rhiz,data = pb)
# TukeyHSD(sw.anova)

shapiro.test(residuals(sw.lm))
# yes, shapiro test says its non-normal, but this test is really sensitive
# and F-tests on balanced data are very robust to to even moderate deviations
# from normality. No need for concern.
hist(resid(sw.lm))
# ^^ looks pretty normal to me.

# qqnorm(residuals(sw.lm))
# qqline(residuals(sw.lm)) redundant

# plot(sw.lm, which = 1)
# plot(sw.lm, which = 2)
```

```{r shoot_weight_diagnostics, echo = F}

# consider using performance::check.model() for diagnostics
library(performance)
library(see)
check_model(sw.lm)
# You have some outliers that have high leverage,
#                    you could try removing them
# The variance inflation factors are somewhat expected, 
#          and check.model(sw.lm2way) shows a reduction in collinearity


summary(sw.lm)
```

```{r log_shoot_weight, echo = F}
# Exploring whether a log transform helps:

# log transform (worth a try)
# pb.log<-pb
# 
# pb.log$sw<-log(pb.log$sw)
# 
# ## this is now the log(log(sw)) vvv
# sw.log.model<-lm(log(sw)~var*rhiz*aph, data=pb)


# **correct way: 
sw.lm.log <- lm(log(sw) ~ var * rhiz * aph, data = pb)

shapiro.test(residuals(sw.lm.log))
hist(resid(sw.lm.log))
```

```{r shoot_weight_log_check_model, echo = F}

check_model(sw.lm.log)
# log transform only improves the PP check.

# sw.glm.model<-glm(sw~var*rhiz*aph,
#                   family = Gamma(link = "log"),
#                   data = pb)
# ^ not necessary

# sanity check to make sure experiment is balanced:
# pb %>% 
#   group_by(var, rhiz, aph) %>% tally() %>% print(n = nrow(.))

```

The simple shoot weight lm is fine, but there isn't a 3-way interaction.
Consider updating your hypothesis.

### Predicted values from shoot weight model:
```{r shoot_weight_fig1, echo = F}

pb_new_dat <- expand.grid(var = levels(pb$var),
                          rhiz = levels(pb$rhiz),
                          aph = levels(pb$aph))

pb_pred <- predict(sw.lm,
                   newdata = pb_new_dat,
                   se.fit = TRUE)

pb_new_dat <- pb_new_dat %>% 
  mutate(fit = pb_pred$fit,
         se = pb_pred$se.fit,
         lwr = fit - (1.96 * se),
         upr = fit + (1.96 * se))

pb_new_dat %>% 
  ggplot(aes(x = aph, y = fit, fill = rhiz)) +
  geom_point(position = position_dodge(width = 0.5),
             size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lwr, ymax = upr, color = rhiz),
                position = position_dodge(width = 0.5),
                width = 0,
                show.legend = F) +
  facet_wrap(~var) +
  theme_minimal() +
  geom_point(data = pb,
             aes(x = aph, y = sw, fill = rhiz),
             position = position_jitterdodge(jitter.width = 0.2,
                                             jitter.height = 0.1,
                                             dodge.width = 0.5),
             shape = 21, alpha = 0.5, color = "black",
             show.legend = F) +
  labs(x = "Aphid treatment",
       y = "Shoot weight (g)",
       fill = "Rhizobia") +
  scale_fill_discrete(labels = c("added",
                               "control")) +
  theme(legend.position = "bottom",
        legend.title.position = "top",
        legend.title = element_text(hjust = 0.5))
  

```

* **Shoot weight by aphid treatment, rhizobia treatment, and wheat variety.** Large points are model estimated predicted values, whiskers are the 95% confidence interval, small points are the raw data.


## Root Weight

```{r Root_weight, echo = F}

# ggplot(pb,aes(x=rw))+
#   geom_histogram(binwidth = 0.1, fill = "cyan", color = "black")+
#   labs(title = "RW", x = "Root Weight", y = "Count")+
#   theme_minimal()

# qqnorm(pb$rw)
# qqline(pb$rw,col="red")
# 
# shapiro.test((pb$rw))

rw.lm<-lm(rw~var*aph*rhiz,data = pb)

check_model(rw.lm) # looks really good

anova(rw.lm) # significant 3-way
summary(rw.lm)

# rw.anova<-aov(rw~var*aph*rhiz,data = pb)
# TukeyHSD(rw.anova)
# I don't think the entire pairwise comparison test is appropriate.
# You have specific hypotheses about how aphids and rhizobia interact
# to affect root weight. So you shouldn't be looking for just any significant comparison, instead, use emmeans and structure your paired contrasts around
# your hypotheses.

```


### Predicted values from root weight model:
```{r root_weight_fig1, echo = F}


root_pred <- predict(rw.lm,
                   newdata = pb_new_dat[1:3],
                   se.fit = TRUE)

root_new_dat <- pb_new_dat[1:3] %>% 
  mutate(fit = root_pred$fit,
         se = root_pred$se.fit,
         lwr = fit - (1.96 * se),
         upr = fit + (1.96 * se))

root_new_dat %>% 
  ggplot(aes(x = aph, y = fit, fill = rhiz)) +
  geom_point(position = position_dodge(width = 0.5),
             size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lwr, ymax = upr, color = rhiz),
                position = position_dodge(width = 0.5),
                width = 0,
                show.legend = F) +
  facet_wrap(~var) +
  theme_minimal() +
  geom_point(data = pb,
             aes(x = aph, y = rw, fill = rhiz),
             position = position_jitterdodge(jitter.width = 0.2,
                                             jitter.height = 0.1,
                                             dodge.width = 0.5),
             shape = 21, alpha = 0.5, color = "black",
             show.legend = F) +
  labs(x = "Aphid treatment",
       y = "Root weight (g)",
       fill = "Rhizobia") +
  scale_fill_discrete(labels = c("added",
                               "control")) +
  theme(legend.position = "bottom",
        legend.title.position = "top",
        legend.title = element_text(hjust = 0.5))
  

```

* **Root weight by aphid treatment, rhizobia treatment, and wheat variety.** Large points are model estimated predicted values, whiskers are the 95% confidence interval, small points are the raw data.

## Aphid nymphs

```{r nymphs, echo = F}

nb$var<-factor(nb$var,levels = c("me","w","l","mi","d","k"),
               labels = c("Melrose","Windham","Lynx",
                          "MiCa","Dint","Klondike"))

nb$aph<-factor(nb$aph,levels = c(1,2),
               labels = c("PEMV-","PEMV+"))

nb$rhiz<-factor(nb$rhiz,levels = c(1,0),
                labels = c("R","NR"))

# ggplot(nb,aes(x=nym))+
#   geom_histogram(binwidth = 0.1, fill = "cyan", color = "black")+
#   labs(title = "nb", x = "Nymph Count", y = "Count")+
#   theme_minimal()
# 
# qqnorm(nb$nym)
# qqline(nb$nym,col="red")

# shapiro.test((nb$nym))

nb.lm<-lm(nym~var*aph*rhiz,data = nb)

check_model(nb.lm) # looks good

anova(nb.lm)
# aphid:rhizobia interaction is weak

summary(nb.lm)

# nb.anova<-aov(nym~var*aph*rhiz,data = nb)
# TukeyHSD(nb.anova)
```

* diagnostics from a lm on nymph counts are fine, but a count model would be more appropriate

```{r nymphs_glm, echo = F}

##################
## an important consideration:

# this model (nymphs) and the model below (nodules) are modeling count data.
# that is, discrete, positive values bound by 0, Inf. Technically, a lm could
# predict negative values as it assumes the data are not bound.
# the model diagnostics look fine, but to be technically correct you should
# use a model that accounts for this type of data.

# consider:

library(glmmTMB)

nymph_mod <- glmmTMB(nym ~ var * aph * rhiz,
                     family = nbinom2(),
                     data = nb)

summary(nymph_mod)
check_model(nymph_mod) # ignore the upper right graph, see below:
check_overdispersion(nymph_mod) # very good sign, good fit overall


```

### Predicted values from nymph model (glm):
```{r nymph_pred_figure, echo = F}

nym_new_dat <- expand.grid(var = levels(nb$var),
                          rhiz = levels(nb$rhiz),
                          aph = levels(nb$aph))


nym_pred <- predict(nymph_mod,
                    newdata = nym_new_dat,
                    se.fit = TRUE,
                    re.form = NA)

nym_new_dat <- nym_new_dat %>% 
  mutate(link.fit = nym_pred$fit,
         link.se = nym_pred$se.fit,
         link.lwr = link.fit - (1.96 * link.se),
         link.upr = link.fit + (1.96 * link.se),
         # put the predictions back on the scale of response variable:
         fit = exp(link.fit),
         lwr = exp(link.lwr),
         upr = exp(link.upr))

nym_new_dat %>% 
  ggplot(aes(x = aph, y = fit, fill = rhiz)) +
  geom_point(position = position_dodge(width = 0.5),
             size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lwr, ymax = upr, color = rhiz),
                position = position_dodge(width = 0.5),
                width = 0,
                show.legend = F) +
  facet_wrap(~var) +
  theme_minimal() +
  geom_point(data = nb,
             aes(x = aph, y = nym, fill = rhiz),
             position = position_jitterdodge(jitter.width = 0.2,
                                             jitter.height = 0.1,
                                             dodge.width = 0.5),
             shape = 21, alpha = 0.5, color = "black",
             show.legend = F) +
  labs(x = "Aphid treatment",
       y = "Aphid nymphs",
       fill = "Rhizobia") +
  scale_fill_discrete(labels = c("added",
                               "control")) +
  theme(legend.position = "bottom",
        legend.title.position = "top",
        legend.title = element_text(hjust = 0.5))


```

* **Count of aphid nymphs by rhizobia and viral treatment.** Large points are model estimated predicted values, whiskers are the 95% confidence interval, small points are the raw data.


## Nodules

```{r nodules, echo = F}

nodb$var<-factor(nodb$var,levels = c("me","w","l","mi","d","k"),
                 labels = c("Melrose","Windham","Lynx",
                            "MiCa","Dint","Klondike"))

nodb$aph<-factor(nodb$aph,levels = c(0,1,2),
                 labels = c("None","PEMV-","PEMV+"))

# nodb$rhiz<-factor(nodb$rhiz,levels = c(1,0),
#                   labels = c("R","NR"))
# your raw data only have one level for rhizobia "1", makes sens as you can't
# have nodules without rhizobia, right???

# correct way:
nodb$rhiz<-NA

# ggplot(nodb,aes(x=nod))+
#   geom_histogram(binwidth = 0.1, fill = "cyan", color = "black")+
#   labs(title = "nodb", x = "Var", y = "Nod")+
#   theme_minimal()

nodb.lm<-lm(nod~aph*var,data = nodb)

check_model(nodb.lm) # again, same as above, model is fine but not
# technically correct. Use a model for count data.

anova(nodb.lm)
# interaction is not significant

summary(nodb.lm)

# nodb.all.lm<-lm(nod~var*aph,data=nodb)
# nodb.anova<-aov(nod~var*aph,data = nodb)
# TukeyHSD(nodb.anova)
```

* again simple LM diagnostics for the nodule model are fine, but a count model would be more appropriate

```{r nodules_glm, echo = F}
###############
# count data model:

nod_mod <- glmmTMB(nod ~ var * aph,
                     family = genpois(), # genpoisson for underdispersion
                     data = nodb)

summary(nod_mod)
# check_model(nod_mod) # wont run with generalized Poisson
# I first ran this with a negative binomial but the model was 
#     actually underdispersed. Otherwise diags looked great.
#     genpois naturally allows for under and overdispersion on a 
#     poisson distribution of counts

check_overdispersion(nod_mod) # dispersion is all good.


```

### Predicted values for nodules (glm):
```{r nod_pred_figure, echo = F}

nod_new_dat <- expand.grid(var = levels(nodb$var),
                          aph = levels(nodb$aph))


nod_pred <- predict(nod_mod,
                    newdata = nod_new_dat,
                    se.fit = TRUE,
                    re.form = NA)

nod_new_dat <- nod_new_dat %>% 
  mutate(link.fit = nod_pred$fit,
         link.se = nod_pred$se.fit,
         link.lwr = link.fit - (1.96 * link.se),
         link.upr = link.fit + (1.96 * link.se),
         # put the predictions back on the scale of response variable:
         fit = exp(link.fit),
         lwr = exp(link.lwr),
         upr = exp(link.upr))

nod_new_dat %>% 
  ggplot(aes(x = aph, y = fit)) +
  geom_point(position = position_dodge(width = 0.5),
             size = 3, shape = 21, color = "black") +
  geom_errorbar(aes(ymin = lwr, ymax = upr),
                position = position_dodge(width = 0.5),
                width = 0,
                show.legend = F) +
  facet_wrap(~var) +
  theme_minimal() +
  geom_jitter(data = nodb,
             aes(x = aph, y = nod),
             shape = 21, alpha = 0.5, color = "black",
             show.legend = F,
             width = 0.2) +
  labs(x = "Aphid treatment",
       y = "Root nodules")

# note the differences in how the variance is calculated between
# a lm and a glm for count data. In the lm, we assume equal variance,
# so the error bars don't vary and the mean is centered. In the glm for count 
# data, the variance is adjusted for each level and the mean isn't centered,
# this is because the data naturally have a left skewed tail.
```

* **Count of nodules by variety and aphid treatment.** Large points are model estimated predicted values, whiskers are the 95% confidence interval, small points are the raw data.

## Contrasts

```{r contrasts, echo = F}
# shoot weight
rhiz_sw.contrast <- emmeans(sw.lm,
                          pairwise ~ rhiz | var * aph)

sw.contrast_df <- as.data.frame(rhiz_sw.contrast$contrasts)

# root weight
rhiz_rw.contrast <- emmeans(rw.lm,
                          pairwise ~ rhiz | var * aph)

rw.contrast_df <- as.data.frame(rhiz_rw.contrast$contrasts)

# nymph counts
rhiz_nc.contrast <- emmeans(nymph_mod, # updated to count model
                          pairwise ~ rhiz | var * aph)

nc.contrast_df <- as.data.frame(rhiz_nc.contrast$contrasts)

###########

# summary tables for figures:

# shoot weight
sw.summary_table <- sw.contrast_df %>%
  rename(Variety = var,
         Aphid = aph,
         Effect = estimate,
         SE = SE,
         P_value = p.value) %>%
  mutate(CI_lower = Effect - 1.96 * SE,
         CI_upper = Effect + 1.96 * SE)

# root  weight
rw.summary_table<-rw.contrast_df %>%
  rename(Variety = var,
         Aphid = aph,
         Effect = estimate,
         SE = SE,
         P_value = p.value) %>%
  mutate(CI_lower = Effect - 1.96 * SE,
         CI_upper = Effect + 1.96 * SE)

# nymphs
nc.summary_table<-nc.contrast_df %>%
  rename(Variety = var,
         Aphid = aph,
         Effect = estimate,
         SE = SE,
         P_value = p.value) %>%
  mutate(lwr = Effect - 1.96 * SE,
         upr = Effect + 1.96 * SE,
         CI_lower = exp(lwr),
         CI_upper = exp(upr),
         Effect = exp(Effect))


```

### Shoot weight contrasts

```{r SW_forest, echo = F}

sw.summary_table %>% 
  ggplot(aes(x=Effect, y=Variety, color=Aphid)) +
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper),
                height = 0,
                position = position_dodge(width = 0.5)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("None" = "black",
                                "PEMV-" = "coral",
                                "PEMV+" = "cyan3"))+
  theme_classic() +
  labs(x = "Rhizobia Effect on Shoot Growth",
       y = "Winter Pea Variety",
       color = "Aphid")

```


### Root weight contrasts

```{r RW_forest, echo =  F}

rw.summary_table %>% 
  ggplot(aes(x=Effect, y=Variety, color=Aphid)) +
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper),
                height = 0,
                position = position_dodge(width = 0.5)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("None" = "black",
                                "PEMV-" = "coral",
                                "PEMV+" = "cyan3")) +
  theme_classic() +
  labs(x = "Rhizobia Effect on Root Growth",
       y = "Winter Pea Variety",
       color = "Aphid")

```


### Nymph contrasts

```{r nymph_forest, echo = F}

nc.summary_table %>% 
  ggplot(aes(x=Effect, y=Variety, color=Aphid)) +
  geom_vline(xintercept = 0, linetype="dashed") +
  geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper),
                width = 0,
                position = position_dodge(width = 0.5)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("PEMV-" = "coral",
                                "PEMV+" = "cyan3"))+
  theme_classic() +
  labs(x = "Rhizobia Effect on Aphid Reproduction",
       y = "Winter Pea Variety",
       color = "Aphid")

```

## Session Information

```{r Session_Info, echo = F, comment = ""}

# Add session information to help with reproduceability
sessionInfo()


```


